---
title: "Vectors and data types"
author:
  - name: Software Carpentry
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
date: 2026-06-08
date-modified: last-modified
engine: knitr
format:
  live-html:
    theme: yeti
    syntax-highlighting: atom-one
    css:
      - styles.css
      - include/webex.css
    include-after-body: include/webex.js
    toc: true
    number-sections: true
    number-depth: 3
    grid:
      body-width: 800px
webr:
  cell-options:
    autorun: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

-----

<br>

## Introduction

### What we'll cover

In this session, you will learn about some of the ways that R stores data.
Specifically, we will cover **vectors** and **data types**.

- **Vectors** are the simplest of the R "data structures",
  object types that R can store data in.
  (In the next session, you'll get introduced to another data structure:
  the data  frame.)

- **Data types** are how R distinguishes between different kinds of data
  like numbers and text.
  We'll talk about the 4 main data types:
  `character`, `integer`, `double`, and `logical`.

### Setting up

To make it easier to keep track of what we do,
we'll write our code in a script:

1. _Open a new R script_:
   Click the **`+`** symbol in the top toolbar, then click `R Script`^[
   Or Click `File` => `New file` => `R Script`.].
   
2. _Save the script_ straight away as `data-structures.R`.
   You can save it anywhere you like,
   but ideally in a folder you made for this workshop.
   
3. If you want a pre-made skeleton with section headers as comments in your script,
   copy-and-paste the following into your script:

   <details><summary>Section headers for your script _(Click to expand)_</summary>
   ```{r, eval=FALSE}
   # 2 - Vectors ------------------------------------------------------------------
   # 2.1 - Single-element vectors (and quoting)

   # Challenge 1:
   # Which of the following option or options will producer? 
   # A) number1 <- "42"
   # B) color <- blue
   # C) animal <- "djfhjkhkjkhCGT"
   # D) number2 <- 10^6

   # 2.2 - Multi-element vectors
   
   # Challenge 2
   # Which elements does the vector birds contain after running the code above?
   # A) cardinal and chickadee
   # B) cardinal, chickadee, and bald eagle
   
   # 2.3 - Vectorization

   # Challenge 3
   # A) Create a vector x with whole numbers 1 through 13,
   #    then print the vector to check your results.
   # B) Subtract 0.5 from each element in x and save the result in vector y,
   #    then print the vector to check your results.

   # 2.4 - Exploring vectors

   # Challenge 4
   # What do you expect `mean(birds)` will return?
   # A) The original vector birds
   # B) Nothing meaningful, maybe an error
   # C) "cardidee"
   
   # 3 - Data types ---------------------------------------------------------------
   # 3.1 - R's main data types

   # 3.2 - A vector can only contain one data type

   # Challenge 5
   # In each line below, what do you think the data type (if any) will be?
   # Try it out and see if you were right.
   # typeof("TRUE")
   # typeof(banana)
   # typeof(c(2, 6, "3"))
   
   # 3.3 - Manual type conversion

   ```
   </details>

## Vectors

A vector in R is essentially **a collection of one or more items**.
Moving forward, we'll call these items "elements".

### Single-element vectors (and quoting)

Vectors can consist of just a single element,
so each of the two lines of code below creates a vector:

```{r}
vector1 <- 8
vector2 <- "panda"
```

In the `"panda"` example, which is a **character string** (_string_ for short):

- `"panda"` constitutes _one element_, not 5 (its number of letters).
- Unlike when dealing with numbers, we have to _quote the string_.^[
  Either double quotes (`"..."`) or single quotes (`'...'`) work,
  but the former are most commonly used by convention.]

Character strings need to be quoted because they are otherwise interpreted as
R objects ---
for example, because `vector1` and `vector2` are objects,
we refer to them without quotes:

```{r}
# Print the values of the two vectors to screen:
vector1
vector2
```

Meanwhile,
the code below doesn't work because there is no _object called `panda`_:

```{r, error=TRUE}
vector_fail <- panda
```

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 1: Quoting**

::: {.callout-tip appearance="simple"}
You can run the code below here on the website
by clicking the **Run Code** button!
:::

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

**Which of the following option or options will produce an error?**
Once you have your answer, feel free to try them out to see what happens.

```{webr}
#| autorun: false
#| caption: ""
number1 <- "42"
number1
```

```{webr}
#| autorun: false
#| caption: ""
color <- blue
color
```

```{webr}
#| autorun: false
#| caption: ""
animal <- "djfhjkhkjkhCGT"
animal
```

```{webr}
#| autorun: false
#| caption: ""
number2 <- 10^6
number2
```

#### {{< fa circle-info >}} Explanation

The only option that will produce an error is the second, `color <- blue`.
The string `"blue"` needs to be quoted,
otherwise R will look for an object called `blue` and return an error if it doesn't exist
(and none exists in this case).

```{r, error=TRUE}
color <- blue
```

<hr style="height:1pt; visibility:hidden;" />

As for the other options:

- Even though `"42"` is a number, it is still a valid string because it is quoted:

  ```{r}
  number1 <- "42"
  number1
  ```

- Even though `"djfhjkhkjkhCGT"` is gibberish, it is still a valid string because it is quoted:

  ```{r}
  animal <- "djfhjkhkjkhCGT"
  animal
  ```

- `10^6` is a valid number in scientific notation, and R will interpret it as 1,000,000:

  ```{r}
  number2 <- 10^6
  number2
  ```

:::
:::

### Multi-element vectors

A common way to make vectors with **multiple elements** is with the
`c` (combine) function:

```{r}
num_vec <- c(2, 6, 3, 41)
num_vec
```

The `c()` function can also **append** elements to an existing vector ---
first we create a vector, then we add an element:

```{r}
# Create a vector with two elements:
birds <- c("cardinal", "chickadee")
birds
```

```{r}
# Add an element to a vector:
c(birds, "bald eagle")
```

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 2: Changing vectors**

::: {.callout-tip appearance="simple"}
Clicking on one of the options below will tell you whether it's correct or not.
Click the Explanation tab for more information.
:::

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

Which elements does the vector `birds` contain after running the code above?

```{r}
#| echo: false
#| results: asis
cat(longmcq(c(
  answer = "A) `cardinal` and `chickadee`",
  'B) `cardinal`, `chickadee`, and `bald eagle`'
)))
```

#### {{< fa circle-info >}} Explanation

The answer is A) `cardinal` and `chickadee`.
The `c()` function does not modify the original vector,
but instead returns a new vector with the additional element.
Since we didn't save the new vector to an object,
the original vector `birds` remains unchanged,
and `bald eagle` is not added to it.

To add the new element to the original vector,
we would have to save the result of `c()` back to `birds`:

```{r}
birds <- c(birds, "bald eagle")
birds
```

:::
:::

<hr style="height:1pt; visibility:hidden;" />

Vectors with **series of numbers** are commonly useful,
and a handy shortcut to make series of whole numbers (integers) is with the
`:` operator:

```{r}
series_vec <- 1:10
series_vec
```

Another example:

```{r}
15:20
```

### Vectorization

Consider the output of these commands:

```{r}
series_vec * 2
```

```{r}
series_vec + 1
```

In the first example, **every individual element in `num_vec` was multiplied by 2**.
Similarly, in the second example, every individual element in `num_vec` was increased by 1.

This behavior,
where an operation is applied to every element of a vector,
is called "vectorization" --- a key feature of the R language!

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 3: Vectorization**

::: {.callout-tip appearance="simple"}
Replace the blanks `______` in the code below with the correct code,
then click **Run Code** to have your answer checked:
:::

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Create a vector `x` with whole numbers 1 through 13,
   then print the vector to check your results.

```{webr}
#| exercise: vectorization1
#| caption: ""
#| autorun: false
#| completion: false
# Create a vector `x` with whole numbers 1 through 13.
x <- ______

# Print vector x
______
```

```{webr}
#| exercise: vectorization1
#| check: true
if (identical(.result, 1:13)) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

```{r}
# Create a vector `x` with whole numbers 1 through 13.
x <- 1:13

# Print vector x
x
```

:::

::: {.panel-tabset}
#### {{< fa list-check >}} Instructions

B) Subtract 0.5 from each element in `x` and save the result in vector `y`,
   then print the vector to check your results.

```{webr}
#| exercise: vectorization2
#| setup: true
x <- 1:13
```

```{webr}
#| exercise: vectorization2
#| caption: ""
#| autorun: false
#| completion: false
# Subtract 0.5 from each element in x and save the result in vector y
y <- ______

# Print vector y
______
```

```{webr}
#| exercise: vectorization2
#| check: true
if (identical(.result, 1:13 - 0.5)) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

```{r}
# Subtract 0.5 from each element in x and save the result in vector y
y <- x - 0.5

# Print vector y
y
```

:::
:::

### Exploring vectors

R has many functions that provide information about vectors and other types of
objects, such as:

- Get the **number of elements** with `length()`:

```{r}
length(series_vec)
```

- See the **first and last few elements**, respectively,
  with `head()` and `tail()`:

  - Print the first 6 elements with `head()`:

    ```{r}
    head(series_vec)
    ```
  
  - Print the last 6 elements with `tail()`:

    ```{r}  
    tail(series_vec)
    ```
  
  - Both `head()` and `tail()` have an argument `n` to specify the number of
    elements:

    ```{r}
    tail(series_vec, n = 2)
    ```

- Get **arithmetic summaries** like `mean()` for vectors with numbers:

  ```{r}
  # mean() will compute the mean (average) across all elements
  mean(series_vec)
  ```

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 4: Prelude to data types**

::: {.callout-tip appearance="simple"}
Clicking on one of the options below will tell you whether it's correct or not.
Click the Explanation tab for more information.
:::

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

What do you expect `mean(birds)` will return?

```{r}
#| echo: false
#| results: asis
cat(longmcq(c(
  "A) The original vector `birds`",
  answer = "B) Nothing meaningful, maybe an error",
  'C) `"cardidee"`'
)))
```

#### {{< fa circle-info >}} Explanation

The answer is B).
Specifically, this code will return a warning along with `NA` (not available):

```{r, error=TRUE}
mean(birds)
```

The `mean()` function is only meant for vectors that consist of numbers,
while `birds` consists of character strings.
We'll dive into this in the next section on data types.

:::
:::

## Data types

### R's main data types

In the exercise above,
we saw that R can't compute the mean of a vector with bird names.
R distinguishes between different kinds of data,
such as character strings and numbers,
using several pre-defined "**data types**".
And its behavior in various operations depends heavily on the data type.

As another example, this fails:

```{r, error=TRUE}
"valerion" * 5
```

We can ask what the data type of something is using the `typeof()` function:

```{r}
typeof("valerion")
```

R has _automatically set_ the data type of `"valerion"` to `character`,
i.e. a (character) string.
And the earlier command failed because R can't perform mathematical functions 
("_binary operator_") on vectors of type `character` ("_non-numeric argument_").

The `character` data type most commonly contains letters,
but anything that is placed between quotes (`"..."`) will be interpreted as this
data type --- even plain numbers:

```{r}
typeof("5")
```

Besides `character`, three other **common data types** are:

- `double` (sometimes referred to as `numeric`) ---
  numbers that can have decimal points:

```{r}
typeof(3.14)
```

- `integer` --- whole numbers only:

```{r}
typeof(1:3)
```

- `logical` (either `TRUE` or `FALSE`, unquoted):

```{r}
typeof(TRUE)
```

### A vector can only contain one data type 

A vector can only **be composed of a single data type**.
As we saw above,
R silently picks the "best-fitting" data type when you create a vector.

::: exercise
#### {{< fa user-edit >}} **Challenge 5: Data types**

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

In each line below, what do you think the data type (if any) will be?
Try it out and see if you were right.

```{webr}
#| autorun: false
#| caption: ""
typeof("TRUE")
```

<hr style="height:1pt; visibility:hidden;" />

```{webr}
#| autorun: false
#| caption: ""
typeof(banana)
```

<hr style="height:1pt; visibility:hidden;" />

```{webr}
#| autorun: false
#| caption: ""
typeof(c(2, 6, "3"))
```

#### {{< fa circle-info >}} Explanation

<hr style="height:1pt; visibility:hidden;" />

1. `"TRUE"` is `character` (and not `logical`) because of the quotes around it:

   ```{r}
   typeof("TRUE")
   ```

-----

2. Recall the earlier example:
   this returns an error because the object `banana` does not exist.
   Any unquoted string (that is not a special keyword like `TRUE` and `FALSE`)
   is interpreted as a reference to an object in R.
   
   ```{r, error=TRUE}
   typeof(banana)
   ```

-----

3. This produces a `character` vector, and we'll talk about why in the next section:
   
   ```{r}
   typeof(c(2, 6, "3"))
   ```

:::

:::

<hr style="height:1pt; visibility:hidden;" />

R's behavior of returning a character vector for `c(2, 6, "3")` in the challenge
above is called **type coercion**.
Because a vector can contain only one data type,
R force-changes any non-conforming elements to the type it has chosen for the vector.

Type coercion can be the source of many surprises,
and is one reason you need to be aware of these data types and R's behavior
around them.

### Manual type conversion

Luckily, you are not just at the mercy of whatever R decides to do automatically,
but can convert vectors using the `as.` group of functions:

::: {.callout-tip appearance="simple"}
Try to use RStudio's auto-complete functionality here: type "`as.`" and then press the <kbd>Tab</kbd> key.
:::

```{r}
as.integer(c("0", "2"))

as.character(c(0, 2))
```

As you may have guessed, though, not all type conversions are possible ---
for example:

```{r}
as.double("kiwi")
```

::: {.callout-note appearance="simple"}
`NA` is R's way of denoting _missing data_ --
see [this bonus section](bonus.qmd#missing-values-na) for more.
:::

That's it for this session!
After lunch, we'll learn to manipulate and summarize data frames using
a real dataset with periodic country-wise statistics such as population size.
