---
title: "Vectors and data types"
author:
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
  - name: Software Carpentry
date: 2026-08-17
date-modified: last-modified
language:
  title-block-published: "Presented on"
  title-block-modified: "Page last modified"
engine: knitr
format:
  live-html:
    theme: yeti
    syntax-highlighting: atom-one
    css:
      - styles.css
      - include/webex.css
    include-after-body: include/webex.js
    toc: true
    number-sections: true
    number-depth: 3
    grid:
      body-width: 800px
webr:
  cell-options:
    autorun: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

-----

<br>

## Introduction

### What we'll cover

In this session, you will learn about some of the ways that R stores data.
Specifically, we will cover **vectors** and **data types**.

- **Vectors** are the simplest of the R "data structures",
  object types that R can store data in.

- **Data types** are how R distinguishes between different kinds of data
  like numbers and text.
  We'll talk about the 4 main data types:
  `character`, `integer`, `double`, and `logical`.

Having a solid understanding of these foundational R concepts will help you avoid
many common pitfalls when working with R.

### Setting up

To make it easier to keep track of what we do,
we'll write our code in a script:

1. _Open a new R script_:
   Click the **`+`** symbol in the top toolbar, then click `R Script`^[
   Or click `File` => `New file` => `R Script`.].
   
2. _Save the script_ straight away as `data-structures.R`.
   You can save it anywhere you like,
   but ideally in a folder you made for this workshop.
   
3. If you want a pre-made skeleton with section headers as comments in your script,
   copy-and-paste the following into your script:

   <details><summary>Section headers for your script _(Click to expand)_</summary>
   ```{r, eval=FALSE}
   # 2 - Vectors ------------------------------------------------------------------
   # 2.1 - Single-item vectors (and quoting)

   # Challenge 1
   # Which of the following options will produce an error?
   # (There may be more than one.)
   # A) number1 <- "42"
   #    number1
   # B) color <- blue
   #    color
   # C) animal <- "djfhjkhkjkhCGT"
   #    animal
   # D) number2 <- 1e6
   #    number2

   # 2.2 - Multi-item vectors
   
   # Challenge 2
   # Which items does the vector birds contain after running the code above?
   # A) cardinal and chickadee
   # B) cardinal, chickadee, and bald eagle
   
   # 2.3 - Vectorization

   # Challenge 3
   # A) Create a vector `x` with whole numbers 1 through 13,
   #    then print the vector to check your results.
   x <- ______

   ______

   # B) Subtract 0.5 from each item in `x` and save the result in vector `y`,
   #    then print the vector to check your results.
   y <- ______

   ______

   # 2.4 - Exploring vectors

   # 3 - Data types ---------------------------------------------------------------
   # 3.1 - R's main data types

   # 3.2 - A vector can only contain one data type

   # Challenge 4
   # In each line below, what do you think the data type will be, if any?
   # Try it out and see if you were right.
   # typeof("TRUE")
   # typeof(banana)
   # typeof(c(2, 6, "3"))
   
   # 3.3 - Manual type conversion
   ```
   </details>

## Vectors

A vector in R is essentially **a collection of one or more items**.

### Single-item vectors (and quoting)

Vectors can consist of just a single item,
so each of the two lines of code below creates a vector:

```{r}
vector1 <- 8
vector2 <- "panda"
```

In the second example, `"panda"` is a **character string** (or simply "string"):

- `"panda"` constitutes _one item_, not 5 (its number of letters).
- Unlike when dealing with numbers, we have to _quote the string_.^[
  Either double quotes (`"..."`) or single quotes (`'...'`) work,
  but the former are most commonly used by convention.]
- This doesn't mean the quotes are part of the string itself:
  they are just a way to tell R that it is a string.

Character strings need to be quoted because without quotes,
R interprets them as **names of objects**.
For example, since `vector1` and `vector2` are objects,
we refer to them without quotes:

```{r}
# Print the values of the two vectors to screen:
vector1
vector2
```

Meanwhile,
the code below doesn't work because there is no object called `panda`:

```{r, error=TRUE}
vector_fail <- panda
```

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 1: Quoting** {#challenge1-quoting}

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

**Which of the following options will produce an error? (There may be more than one.)**

First think about it, _then_ click the **Run Code** button to check your answer.
Afterwards, you can also click the **Explanation** tab to see why.

```{webr}
#| autorun: false
#| caption: ""
number1 <- "42"
number1
```

```{webr}
#| autorun: false
#| caption: ""
color <- blue
color
```

```{webr}
#| autorun: false
#| caption: ""
animal <- "djfhjkhkjkhCGT"
animal
```

```{webr}
#| autorun: false
#| caption: ""
number2 <- 1e6
number2
```

#### {{< fa circle-info >}} Explanation

The only option that will produce an error is the second, `color <- blue`.
Here, `blue` needs to be quoted as `"blue"`.
Without the quotes, R looks for an *object* called `blue`,
and returns an error because no such object exists.

```{r, error=TRUE}
color <- blue
```

<hr style="height:1pt; visibility:hidden;" />

As for the other options:

- Even though `"42"` is a number, it is still a valid string because it is quoted:

  ```{r}
  number1 <- "42"
  number1
  ```

- Even though `"djfhjkhkjkhCGT"` is gibberish, it is still a valid string because it is quoted:

  ```{r}
  animal <- "djfhjkhkjkhCGT"
  animal
  ```

- `1e6` is a valid number in scientific notation, and R will interpret it as 1 million:

  ```{r}
  number2 <- 1e6
  number2
  ```

:::
:::

### Multi-item vectors

A common way to make vectors with **multiple items** is with the
`c` (combine) function:

```{r}
num_vec <- c(2, 6, 3, 41)
num_vec
```

The `c()` function can also **add** items to a vector you already have.
Below, we first create a vector,
and then use that vector as an argument to the `c()` function:

```{r}
birds <- c("cardinal", "chickadee")
birds
```

```{r}
c(birds, "bald eagle")
```

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 2: Changing vectors** {#challenge2-changing-vectors}

::: {.callout-tip appearance="simple"}
Click on one of the options below to make your selection,
and the page will tell you whether it is correct.
You can then also click the Explanation tab for more information.
:::

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

Which items does the vector `birds` contain after running the code above?

```{r}
#| echo: false
#| results: asis
cat(longmcq(c(
  answer = "A) `cardinal` and `chickadee`",
  'B) `cardinal`, `chickadee`, and `bald eagle`'
)))
```

#### {{< fa circle-info >}} Explanation

The answer is A) `cardinal` and `chickadee`.

The `c()` function **does not modify the original vector,**
**but instead returns a new vector with the additional item.**

Since we didn't save the new vector to an object,
the original vector `birds` remains unchanged,
and `bald eagle` is not added to it.

To add the new item to the original vector,
we would save the result of `c()` back to `birds`:

```{r}
birds <- c(birds, "bald eagle")
birds
```

:::
:::

<hr style="height:1pt; visibility:hidden;" />

Vectors containing a **series of numbers** are often useful,
and the `:` operator is a handy shortcut for making series of whole numbers
(integers):

```{r}
series_vec <- 1:10
series_vec
```

Another example:

```{r}
15:20
```

### Vectorization

Consider the output of these commands:

```{r}
series_vec * 2
```

```{r}
series_vec + 1
```

In the first example, **every individual item in `series_vec` was multiplied by 2**.
Similarly, in the second example, every individual item in `series_vec` was
increased by 1.

This behavior,
where an operation is automatically applied to every item of a vector,
is called "vectorization" --- a key feature of the R language!

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} **Challenge 3: Vectorization** {#challenge3-vectorization}

Replace the blanks `______` in the code below with the correct code,
then click **Run Code** to have your answer checked:

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Create a vector `x` with whole numbers 1 through 13,
   then print the vector to check your results.

```{webr}
#| exercise: vectorization1
#| caption: ""
#| autorun: false
#| completion: false
# Create a vector `x` with whole numbers 1 through 13.
x <- ______

# Print vector x to check your results.
______
```

```{webr}
#| exercise: vectorization1
#| check: true
if (identical(.result, 1:13)) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

```{r}
# Create a vector `x` with whole numbers 1 through 13.
x <- 1:13

# Print vector x to check your results.
x
```

:::

<hr style="height:1pt; visibility:hidden;" />

::: {.panel-tabset}
#### {{< fa list-check >}} Instructions

B) Subtract 0.5 from each item in `x` and save the result in vector `y`,
   then print the vector to check your results.

```{webr}
#| exercise: vectorization2
#| setup: true
x <- 1:13
```

```{webr}
#| exercise: vectorization2
#| caption: ""
#| autorun: false
#| completion: false
# Subtract 0.5 from each item in x and save the result in vector y.
y <- ______

# Print vector y to check your results.
______
```

```{webr}
#| exercise: vectorization2
#| check: true
if (identical(.result, 1:13 - 0.5)) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

```{r}
# Subtract 0.5 from each item in x and save the result in vector y
y <- x - 0.5

# Print vector y
y
```

:::
:::

### Exploring vectors

R has many functions that provide information about vectors and other types of
objects, such as:

- Get the **number of items** with `length()`:

  ```{r}
  length(series_vec)
  ```

- See the **first few items** with `head()` (or the last few with `tail()`):

  - By default, it prints the first 6 items:

    ```{r}
    head(series_vec)
    ```

  - You can specify the number of items with the `n` argument:

    ```{r}
    head(series_vec, n = 2)
    ```

- Get **arithmetic summaries** like `mean()` for vectors with numbers:

  ```{r}
  # mean() will compute the mean (average) across all items
  mean(series_vec)
  ```

## Data types

### R's main data types

R distinguishes between different kinds of data,
such as character strings and numbers,
with several pre-defined "**data types**".
Its behavior in various operations depends heavily on the data type.

For example, R refuses to perform mathematical operations on character strings,
so the code below will return an error:

```{r, error=TRUE}
"valerion" * 5
```

We can ask what the data type of something is using the `typeof()` function:

```{r}
typeof("valerion")
```

R has automatically set the data type of `"valerion"` to `character`,
i.e. a character string.

Note that while the `character` data type most commonly contains letters,
anything that is placed between quotes (`"..."`) will be interpreted as this
data type --- even plain numbers:

```{r}
typeof("5")
```

Besides `character`, three other **common data types** are:

- `double` (sometimes referred to as `numeric`) ---
  numbers that can have decimal points:

  ```{r}
  typeof(3.14)
  ```

- `integer` --- whole numbers only:

  ```{r}
  typeof(1:3)
  ```

- `logical` (either `TRUE` or `FALSE`, unquoted!):

  ```{r}
  typeof(TRUE)
  ```

### A vector can only contain one data type

A vector can only **be composed of a single data type**.
As we saw above,
R silently picks the "best-fitting" data type when you create a vector.

::: exercise
#### {{< fa user-edit >}} **Challenge 4: Data types** {#challenge4-data-types}

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

In each line below, what do you think the data type will be, if any?
Try it out and see if you were right.

```{webr}
#| autorun: false
#| caption: ""
typeof("TRUE")
```

<hr style="height:1pt; visibility:hidden;" />

```{webr}
#| autorun: false
#| caption: ""
typeof(banana)
```

<hr style="height:1pt; visibility:hidden;" />

```{webr}
#| autorun: false
#| caption: ""
typeof(c(2, 6, "3"))
```

#### {{< fa circle-info >}} Explanation

<hr style="height:1pt; visibility:hidden;" />

1. `"TRUE"` is `character` (and not `logical`) because of the quotes around it:

   ```{r}
   typeof("TRUE")
   ```

-----

2. Recall the earlier example:
   this returns an error because the object `banana` does not exist.
   Any unquoted string (that is not a special keyword like `TRUE` and `FALSE`)
   is interpreted as a reference to an object in R.
   
   ```{r, error=TRUE}
   typeof(banana)
   ```

-----

3. This produces a `character` vector, and we'll talk about why right below
   this challenge:
   
   ```{r}
   typeof(c(2, 6, "3"))
   ```

:::

:::

<hr style="height:1pt; visibility:hidden;" />

R's behavior of returning a character vector for `c(2, 6, "3")` in the challenge
above is called **type coercion**.
More generally, R forcibly converts any non-conforming items to the type it
has chosen for a vector.

::: {.callout-note appearance="simple"}
Type coercion can be the source of many surprises,
and is one reason you need to be aware of data types and R's behavior
around them.
:::

### Manual type conversion

You are not at the mercy of whatever R decides to do automatically:
you can also convert vectors yourself, using the `as.` group of functions:

::: {.callout-tip appearance="simple"}
Try to use RStudio's auto-complete functionality here: type "`as.`" and then press the <kbd>Tab</kbd> key.
:::

```{r}
as.integer(c("0", "2"))

as.character(c(0, 2))
```

As you may have guessed, though, not all type conversions are possible ---
for example:

```{r}
as.double("kiwi")
```

::: {.callout-note appearance="simple"}
`NA` is R's way of denoting _missing data_ --
see [this bonus section](bonus.qmd#missing-values-na){target="_blank"} for more.
:::

That's it for this session!
After lunch, we'll learn to manipulate and summarize data frames using
a real dataset with statistics such as population size for many countries
over time.

<br>
