---
title: "Data wrangling with _dplyr_"
author:
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
  - name: Software Carpentry
date: 2026-06-17
date-modified: last-modified
language:
  title-block-published: "Presented on"
  title-block-modified: "Page last modified"
engine: knitr
format:
  live-html:
    theme: yeti
    syntax-highlighting: atom-one
    css:
      - styles.css
      - include/webex.css
    include-after-body: include/webex.js
    toc: true
    number-sections: true
    number-depth: 3
    grid:
      body-width: 800px
webr:
  cell-options:
    autorun: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| label: setup
#| include: false
#| context: setup
library(gapminder)
library(dplyr)
```

-------

<br>

## Introduction

### What we'll cover

Many of the most common data analysis tasks involve working with **tabular data**,
i.e. data with rows and columns. In this session,
you'll learn to work with tabular data in R using the _dplyr_ package.

For example, you'll see how to subset, manipulate and summarize a dataset ---
the kinds of tasks commonly referred to as "data wrangling".
Such data processing is often essential before you can move on to
data visualization (next session of this workshop!) or statistical analyses.

### Setting up

Use a new script for this session, much like we did for the previous one:

1. _Open a new R script_ (Click the **`+`** symbol in the toolbar at the top, then click `R Script`)^[
   Or click `File` => `New file` => `R Script`.].
   
2. _Save the script_ straight away as `data-wrangling.R` --
   you can save it anywhere you like, though it is probably best to save it in a
   folder specifically for this workshop.
   
3. If you want the section headers as comments in your script,
   like in the script I am showing you in the live session,
   then copy-and-paste the following into your script:

   <details><summary>Section headers for your script _(Click to expand)_</summary>
   ```{r, eval=FALSE}
   # 1 - Introduction -------------------------------------------------------------
   # 1.4 - Loading the tidyverse

   # 2 - The gapminder dataset ----------------------------------------------------

   # Challenge 1
   # The country column repeats the same country many times.
   # Scroll down in the View() tab: what makes each row unique?

   # 3 - arrange() ----------------------------------------------------------------

   # Challenge 2
   # Which line of code would you use to find the row with the highest
   # per-capita GDP?
   # A) arrange(.data = gapminder, desc(gdpPercap), year, country)
   # B) arrange(.data = gapminder, country, year, gdpPercap)
   # C) arrange(.data = gapminder, desc(gdpPercap))
   # D) arrange(.data = gapminder, gdpPercap)

   # 4 - filter() -----------------------------------------------------------------
   # 4.1 - Filter based on one condition

   # 4.2 - Filter based on multiple conditions

   # 5 - The pipe (|>) ------------------------------------------------------------

   # Challenge 3
   # The pipeline below is supposed to find the 3 countries with the lowest
   # GDP per capita in 1997, sorted from lowest to highest.
   # It has two errors. Find and fix them:
   gapminder
   filter(year = 1997) |>
     arrange(gdpPercap) |>
     head(n = 3)

   # Challenge 4
   # Using filter(), arrange(), and the pipe (|>), answer the following:
   # A) Which African country had the highest life expectancy in 2002?

   # B) What was the smallest population recorded in Asia across all years?

   # 6 - mutate() -----------------------------------------------------------------

   # Challenge 5
   # Use mutate() to create a new column gdp_billion that has the absolute GDP
   # (i.e., not relative to population size) in units of billions.
   gapminder |>
     mutate(______)

   # 7 - summarize() --------------------------------------------------------------

   # Challenge 6
   # A) Calculate the average life expectancy for each country and store the
   #    result in a new data frame.
   gapminder |>
     ______ |>
     ______

   # B) Use the data frame you just created to find out which country has the
   #    longest average life expectancy and which has the shortest.
   ```

   </details>

### The _dplyr_ package and the tidyverse

One of R's most powerful features is its ability to deal with **tabular data**.
That is, data with rows and columns like those familiar from Excel spreadsheets
and so on. 
R stores tabular data in a data structure called a "**data frame**".

The _dplyr_ package provides
useful functions for **manipulating data in data frames**.
In this session, we'll cover the following commonly used _dplyr_ functions:

- `arrange()` to change the order of rows (i.e., to sort a data frame)
- `filter()` to keep only a subset of rows
- `mutate()` to manipulate columns and create new columns
- `summarize()` to compute data summaries across rows (if we have time)

::: {.callout-note appearance="simple"}
_dplyr_ has many more functions, and we encourage you to explore these on
your own: [bonus material on this site](./bonus.qmd#bonus-material-for-session-3){target="_blank"} /
[dplyr documentation](https://dplyr.tidyverse.org/articles/dplyr.html){target="_blank"}. 
:::

_dplyr_, in turn, belongs to the ["tidyverse"](https://www.tidyverse.org){target="_blank"},
a **family of R packages for data science**.
Another key tidyverse package we'll cover in today's workshop is _ggplot2_
for making plots.

### Loading the tidyverse

You should have the tidyverse already _installed_,
but in every new R session, you need to _load_ it before you can use it.
Let's do that now:

```{r}
library(tidyverse)
```

The output tells you which packages have been loaded as part of the tidyverse.
(For now, you don't have to pay attention to the
["Conflicts" section](./bonus.qmd#function-name-conflicts){target="_blank"}.)

::: {.callout-warning collapse="true"}
### Didn't install those packages yet? _(Click to expand)_

```{r, eval=FALSE}
install.packages("tidyverse")
install.packages("gapminder")
```

:::

## The `gapminder` dataset

In this session and the next one, we will work with the `gapminder` dataset.
This dataset contains
**statistics such as population size and life expectancy for 142 countries,**
**at five-year intervals from 1952 to 2007.**

It is available in a package of the same name.
Like with the tidyverse,
you should have it installed but do, as always, need to load it:

```{r}
# (Unlike with the tidyverse, no output is expected when you load gapminder)
library(gapminder)
```

Take a look at the `gapminder` data frame that is now available to you:

```{r}
gapminder
```

::: {.callout-note appearance="simple"}
Above, the `gapminder` data frame is referred to as a *"tibble"*,
which is the tidyverse's version of a data frame, with a few small differences^[
The main difference is the nicer default printing behavior of tibbles:
e.g. the data types of columns are shown,
and only a limited number of rows are printed.].
:::

Only the first 10 rows are printed, and the dataset has the following columns:

| column name | data type           | meaning                                      |
| ----------- | ------------------- | -------------------------------------------- |
| `country`   | `factor` (`<fct>`)  | Country                                      |
| `continent` | `factor` (`<fct>`)  | Continent that the country is in             |
| `year`      | `integer` (`<int>`) | Focal year for the stats in the next columns |
| `lifeExp`   | `double` (`<dbl>`)  | Mean life expectancy in years                |
| `pop`       | `integer` (`<int>`) | Population size                              |
| `gdpPercap` | `double` (`<dbl>`)  | Per-capita GDP                               |

: {.striped tbl-colwidths="[18,22,56]"}

<hr style="height:1pt; visibility:hidden;" />

::: {.callout-tip appearance="simple"}
Two columns are of type `factor`,
[an alternative to](./bonus.qmd#factors){target="_blank"}
`character` commonly used for categorical data.
:::

You can also use the `View()` function to look at the data frame.
This will open a new tab in your editor pane with a spreadsheet-like look and feel:

```{r, eval=FALSE}
View(gapminder)
# (Should display the dataset in an editor pane tab)
```

::: exercise
#### {{< fa user-edit >}} **Challenge 1: Explore the `gapminder` dataset** {#challenge1-explore-gapminder}

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

The `country` column repeats the same country many times.
Scroll down in the `View()` tab: what makes each row unique?

#### {{< fa circle-check >}} Solution

Each row is one **country in one year**.
Afghanistan, for example, takes up 12 rows:
one for each year from 1952 to 2007, in five-year steps.

:::
:::

## `arrange()`

The `arrange()` function is like sorting functionality in Excel:
it changes the order of rows based on the values in one or more columns.
Entire rows are always moved together:
it never reorders the values within a single column on its own,
because that would scramble the data.

`gapminder` is currently first sorted alphabetically by `country`,
and next by `year`.
You may want to sort, for example, by population size instead,
and can do so with `arrange()` as follows:

```{r}
gm_arranged <- arrange(.data = gapminder, pop)
```

Before we take a look at the result,
let's break down the syntax of the `arrange()` function:

- The **first argument** is the data frame to be sorted,
  which we specify with the `.data` argument.
- The **second argument** is the column to sort by,
  which we specify without an argument name.

Like all _dplyr_ functions, `arrange()` outputs a new data frame,
and **does not modify the input data frame.**
In this case, we chose to assign the output to a new object called `gm_arranged`.

<details><summary>[{{< fa comments >}}]{style="color: darkgreen;"} What would have happened if we had assigned the output to `gapminder`?</summary>
In that case, the original data frame would have been overwritten with the sorted version.
</details>

Finally, let's look at the result of the sorting:

```{r}
gm_arranged
```

Sorting helps you find observations with the smallest or largest values
for a certain column: above, we see that the dataset's smallest population size
is from _Sao Tome and Principe_ in 1952.

Default sorting is from **small to large**, as seen above.
To sort in reverse order, use the `desc()` (descending) helper function:

```{r}
arrange(.data = gapminder, desc(pop))
```

::: {.callout-tip appearance='simple'}
Here, we didn't assign the output to a new object, so it was just printed to screen.
We'll keep doing that for the remainder of the session, 
so we can more easily see the results of our operations.
:::

Finally, you may want to sort by multiple columns,
where ties in the first column are broken by a second column (and so on).
Do this by simply listing the columns in the appropriate order:

```{r}
# Sort first by continent, then by country:
arrange(.data = gapminder, continent, country)
```

Pay attention to the syntax:
additional columns to sort by are additional arguments to the function,
not part of the same argument with `c()`.
Other _dplyr_ functions work the same way.

:::exercise
#### {{< fa user-edit >}} **Challenge 2: `arrange()`** {#challenge2-arrange}

One observation (row) in the `gapminder` dataset has the highest per-capita GDP.
Which line of code would you use to find out which row that is?

Try it, and report which row (which country and year) you found.

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

```{r}
#| echo: false
#| results: asis
cat(longmcq(c(
  'A) `arrange(.data = gapminder, desc(gdpPercap), year, country)`',
  'B)`arrange(.data = gapminder, country, year, gdpPercap)`',
  answer = 'C) `arrange(.data = gapminder, desc(gdpPercap))`',
  'D) `arrange(.data = gapminder, gdpPercap)`'
)))
```

#### {{< fa circle-info >}} Explanation

The best answer is `C`.
While both `A` and `C` give you the row with the highest per-capita GDP,
`A` will also sort the entire dataset by year and country,
which is not necessary to answer the question.

Most importantly, you want to sort by `gdpPercap` in descending order,
so the row with the highest value ends up at the top:

```{r}
arrange(.data = gapminder, desc(gdpPercap))
```

Kuwait in 1957 had the highest per-capita GDP in the dataset, at $113,523.
:::
:::

## `filter()`

The `filter()` function outputs only those rows that satisfy one or more conditions.
It is similar to filtering functionality in Excel.

### Filter based on one condition

This first `filter()` example outputs only rows for which the
life expectancy exceeds 80 years
(remember, each row represents a country in a given year):

```{r}
filter(.data = gapminder, lifeExp > 80)
```

<details><summary>[{{< fa comments >}}]{style="color: darkgreen;"} How many rows were output? _(Click to see the answer)_</summary>
21 rows were output. This is most easily seen in the first line of the output
(`21 x 6`, i.e. 21 rows and 6 columns).

</details>

As the example above demonstrated,
`filter()` **outputs rows that satisfy the condition(s)** you specify.
These conditions don't have to be based on numeric comparisons:

```{r}
# Only keep rows where the value in the 'continent' column is 'Europe':
filter(.data = gapminder, continent == "Europe")
```

::: {.callout-warning appearance='simple'}
Remember to use **two** equals signs `==` to test for equality!
:::

### Filter based on multiple conditions

It's also possible to filter based on multiple conditions.
For example, you may want to see which countries in Asia had a life expectancy
greater than 80 years:

```{r}
filter(.data = gapminder, continent == "Asia", lifeExp > 80)
```

As in the example above, multiple conditions are by default combined
using a Boolean ***AND***.
In other words, in a given row, _each_ condition must be met to output the row.

::: {.callout-note collapse="true"}
#### What about a Boolean ***OR***? _(Click to expand)_

If you want to combine conditions using a Boolean ***OR***,
where only one of the conditions needs to be met,
use a `|` (vertical bar) between the conditions:

```{r}
# Keep rows with a high life expectancy and/or a high per-capita GDP:
filter(.data = gapminder, lifeExp > 80 | gdpPercap > 100000)
```
:::

## The pipe (`|>`)

The examples so far applied a single _dplyr_ function to a data frame.
But in practice, it's common to use several consecutive _dplyr_ functions to
wrangle a data frame into the format you want.

For example, you may want to filter rows and then sort the result.
You could do that as follows:

```{r}
gm_filt <- filter(.data = gapminder, lifeExp < 50)

arrange(.data = gm_filt, desc(pop))
```

And you could go on like this,
successively creating new objects you use for the next step.
But there is a more elegant way of doing this!

You can directly send ("pipe") output from one function into the next function
with the **pipe operator** `|>`,
which is a vertical bar <kbd>|</kbd> followed by a greater-than sign <kbd>></kbd>.

Let's start by seeing a rephrasing of the code above, now using pipes:

```{r}
gapminder |>
  filter(lifeExp < 50) |>
  arrange(desc(pop))
```

What happened here? We took the `gapminder` data frame,
sent ("piped") it into the `filter()` function,
whose output in turn was piped into the `arrange()` function.
You can think of the pipe as **"then"**:
take `gapminder`, _then_ filter, _then_ arrange.

When using the pipe,
you no longer specify the input data frame with the `.data` argument,
because the function now gets its input data via the pipe^[
Specifically, the input goes to the function's first argument by default.].

::: {.callout-tip}
#### Pipe tips
Using pipes involves less typing and, above all, is more readable than using
successive assignments^[
Using pipes also avoids cluttering your environment with intermediate
objects, which saves computer memory.
].

For code readability, 
it is good practice to always start a new line after a pipe `|>`,
and to keep the subsequent line(s) _indented_ as RStudio will automatically do.
:::

:::exercise
#### {{< fa user-edit >}} **Challenge 3: Find the mistakes** {#challenge3-find-mistakes}

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

The below "pipeline" is supposed to find the 3 countries with the lowest GDP per
capita in 1997, sorted from lowest to highest.
It has two errors. Find and fix them:

```{webr}
#| autorun: false
#| caption: ""
#| exercise: find_errors
gapminder
filter(year = 1997) |>
  arrange(gdpPercap) |>
  head(n = 3)
```

```{webr}
#| exercise: find_errors
#| check: true
if (
  identical(
    .result,
    gapminder |> filter(year == 1997) |> arrange(gdpPercap) |> head(n = 3)
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

**The two errors are:**

- Line 1: Missing pipe `|>` after gapminder
- Line 2: Should use `==` (not `=`) to test for equality

**Corrected code:**

```{r}
gapminder |>
  filter(year == 1997) |>
  arrange(gdpPercap) |>
  head(n = 3)
```

:::
:::

:::exercise
#### {{< fa user-edit >}} **Challenge 4: Find the extremes** {#challenge4-find-extremes}

Using `filter()`, `arrange()`, and the pipe (`|>`),
answer the following questions about the gapminder dataset:

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Which African country had the highest life expectancy in 2002?

#### {{< fa circle-check >}} Solution

We'll use the same `head()` trick as in the exercise above
to see _only_ the top country, but this is not strictly necessary:
   
```{r}
gapminder |>
  filter(continent == "Africa", year == 2002) |>
  arrange(desc(lifeExp)) |>
  head(n = 1)
```

:::
::: {.panel-tabset}

<hr style="height:1pt; visibility:hidden;" />

#### {{< fa list-check >}} Instructions

B) What was the smallest population recorded in Asia across all years?

#### {{< fa circle-check >}} Solution
   
```{r}
gapminder |>
  filter(continent == "Asia") |>
  arrange(pop) |>
  head(n = 1)
```
  
:::
:::

## `mutate()`

So far, we've focused on functions that subset and reorganize data frames.
But we haven't yet seen how to _change the data_ or _compute derived data_.
This can be done with the `mutate()` function.

For example, to create a new column with population sizes in millions:

```{r}
# Create a new column 'pop_million' by dividing 'pop' by a million:
gapminder |>
  mutate(pop_million = pop / 10^6)
```

To _modify_ an existing column rather than adding a new one,
simply "assign back to the same name":

```{r}
# Change the unit of the 'pop' column:
gapminder |>
  mutate(pop = pop / 10^6)
```

:::exercise
#### {{< fa user-edit >}} **Challenge 5: Absolute GDP** {#challenge5-absolute-gdp}

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

Use `mutate()` to create a new column `gdp_billion` that has the
**absolute** GDP (i.e., not relative to population size)
in **units of billions**.

```{webr}
#| autorun: false
#| caption: ""
#| exercise: abs_gdp
gapminder |>
  mutate(______)
```

```{webr}
#| exercise: abs_gdp
#| check: true
if (
  identical(
    .result,
    gapminder |> mutate(gdp_billion = gdpPercap * pop / 10^9)
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa lightbulb >}} Hints

```{r, eval=FALSE}
gapminder |>
  mutate(gdp_billion = gdpPercap * ______ / ______)
```

#### {{< fa circle-check >}} Solution

You need to do two things, which can be combined into a single line:

1. Make the GDP absolute by multiplying by the population size: `gdpPercap * pop`
2. Change the unit of the absolute GDP to billions by dividing by a billion:
   ` / 10^9`

```{r}
gapminder |>
  mutate(gdp_billion = gdpPercap * pop / 10^9)
```

:::
:::

## `summarize()` _(If we have time)_

The final _dplyr_ function we'll cover is `summarize()`,
which computes **summaries** of your data **across rows**.
For example, to calculate the mean GDP across the entire dataset:

```{r}
# The syntax is similar to 'mutate': <new-column> = <operation>
gapminder |>
  summarize(mean_gdp = mean(gdpPercap))
```

The output is still a data frame, but unlike with the previous _dplyr_
functions, its rows are not the rows of the input:
`summarize()` "collapses" the data down, here to a single number.

`summarize()` becomes really powerful in combination with `group_by()`,
which lets you compute **groupwise** stats.
For example, to get the mean GDP separately for each continent:

```{r}
gapminder |>
  group_by(continent) |>
  summarize(mean_gdp = mean(gdpPercap))
```

`group_by()` implicitly splits a data frame into groups of rows:
here, one group for observations from each continent.
After that, operations like in `summarize()` will happen separately for each group,
which is how we ended up with per-continent means.

:::exercise
#### {{< fa user-edit >}} **Challenge 6: Mean life expectancy** {#challenge6-mean-life-expectancy}

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Calculate the average life expectancy for each country and store the result in
   a new data frame.

```{webr}
#| autorun: false
#| caption: ""
#| exercise: life_exp
gapminder |>
  ______ |>
  ______
```

```{webr}
#| exercise: life_exp
#| check: true
if (
  identical(
    .result,
    lifeExp_bycountry <- gapminder |> group_by(country) |> summarize(mean_lifeExp = mean(lifeExp))
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa lightbulb >}} Hints

```{r, eval=FALSE}
lifeExp_bycountry <- gapminder |>
  group_by(______) |>
  summarize(mean_lifeExp = ______)
```

#### {{< fa circle-check >}} Solution

First, create a data frame with the mean life expectancy by country:

```{r}
lifeExp_bycountry <- gapminder |>
  group_by(country) |>
  summarize(mean_lifeExp = mean(lifeExp))
```

:::
::: {.panel-tabset}

<hr style="height:1pt; visibility:hidden;" />

#### {{< fa list-check >}} Instructions

B) Use the data frame you just created to find out which country has the longest average
   life expectancy and which has the shortest.

#### {{< fa circle-check >}} Solution

Sort the data frame twice, once in each direction, to see the countries
with the shortest and the longest life expectancy.
You could optionally pipe into `head()` to only see the top `n`, here top 1:

```{r}
lifeExp_bycountry |>
  arrange(mean_lifeExp) |>
  head(n = 1)

lifeExp_bycountry |>
  arrange(desc(mean_lifeExp)) |>
  head(n = 1)
```

So, **Sierra Leone** has the shortest average life expectancy (36.8 years),
and **Iceland** has the longest average life expectancy (76.5 years).
:::
:::

<br>
