---
title: "Data wrangling with _dplyr_"
author:
  - name: Software Carpentry
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
date: 2026-06-08
date-modified: last-modified
engine: knitr
format:
  live-html:
    theme: yeti
    syntax-highlighting: atom-one
    css:
      - styles.css
      - include/webex.css
    include-after-body: include/webex.js
    toc: true
    number-sections: true
    number-depth: 3
    grid:
      body-width: 800px
webr:
  cell-options:
    autorun: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| label: setup
#| include: false
#| context: setup
library(gapminder)
library(dplyr)
```

-------

<br>

## Introduction

### What we'll cover

Many of the most common data analysis tasks involve working with **tabular data**,
i.e. data with rows and columns. In this session,
you'll learn to work with tabular data in R using the _dplyr_ package.

For example, you'll see how to subset, manipulate and summarize a dataset,
the kinds of tasks commonly referred to as "data wrangling".
Such data processing is often essential before you can move on to say,
data visualization (next session of this workshop) or statistical analyses.

### Setting up

Use a new script for this session, much like we did for the previous one:

1. _Open a new R script_ (Click the **`+`** symbol in toolbar at the top, then click `R Script`)^[
   Or Click `File` => `New file` => `R Script`.].
   
2. _Save the script_ straight away as `data-wrangling.R` --
   you can save it anywhere you like, though it is probably best to save it in a
   folder specifically for this workshop.
   
3. If you want the section headers as comments in your script,
   like in the script I am showing you in the live session,
   then copy-and-paste the following into your script:

   <details><summary>Section headers for your script _(Click to expand)_</summary>
   ```{r}

   # TODO - UPDATE

   # 1 - Introduction ----------------------------------------------------
   # 1.4 - Loading the tidyverse

   # 2 - The gapminder dataset  ------------------------------------------
   
   # Challenge 1

   # 3 - arrange() -------------------------------------------------------

   # Challenge 2

   # 4 - filter() --------------------------------------------------------
   # 4.1 - Filter based on one condition

   # 4.2 - Filter based on multiple conditions

   # 5 - The pipe --------------------------------------------------------

   # Challenge 3

   # Challenge 4

   # 6 - mutate() --------------------------------------------------------

   # Challenge 5

   # 7 - summarize() -----------------------------------------------------

   # Challenge 6
   ```

   </details>

### The _dplyr_ package and the tidyverse

One of R's most powerful features is its ability to deal with **tabular data**:
data with rows and columns like you are familiar with from Excel spreadsheets
and so on --- for example:

<hr style="height:1pt; visibility:hidden;" />

```{r, echo=FALSE}
head(data.frame(gapminder::gapminder))
```

In R, tabular data is stored in a data structure called "**data frame**".
The [_dplyr_](https://cran.r-project.org/package=dplyr) package provides very
useful functions for **manipulating data in data frames**.
In this session, we'll cover these commonly used _dplyr_ functions:

- `arrange()` to change the order of rows (i.e., to sort a data frame)
- `filter()` to keep only a subset of rows
- `mutate()` to manipulate columns and create new columns
- `summarize()` to compute data summaries across rows

_dplyr_ has many more functions.
Some of these are more advanced than those we'll cover today,
while others are conceptually simpler
(e.g. `rename()` to change column names and `select()` to select columns),
and we encourage you to [explore](#bonus-material-for-self-study) these
[on your own](https://dplyr.tidyverse.org/articles/dplyr.html). 

_dplyr_ belongs to a **family of R packages for data science called the**
["tidyverse"](https://www.tidyverse.org).**
Another tidyverse package we'll cover in today's workshop is _ggplot2_
for making plots.

### Loading the tidyverse

All core tidyverse packages can be installed and loaded with a single command.
Since you should already have _installed_ the tidyverse^[
If not: run `install.packages("tidyverse")` now.
], you only need to _load_ it, which you do as follows:

```{r}
library(tidyverse)
```

The output tells you which packages have been loaded as part of the tidyverse.
(For now, you don't have to pay attention to or worry about the
["Conflicts" section](#function-name-conflicts).)

## The `gapminder` dataset

In this session and the next one on data visualization,
we will work with the `gapminder` dataset,
which contains statistics such as population size for different countries across
five-year intervals.

This dataset is available in a package of the same name,
which you already should have installed as well^[
If not: run `install.packages("gapminder")` now.
]. To load the package into your current R session:

```{r}
# (Unlike with the tidyverse, no output is expected when you load gapminder)
library(gapminder)
```

Take a look at the `gapminder` data frame that is now available to you:

```{r}
gapminder
```

The first 10 rows are printed, with the following columns:

| column name | data type | meaning |
|---|---|------|
| `country` | `factor` (`<fct>`) | Country |
| `continent` | `factor` (`<fct>`) | Continent that the country is in |
| `year` | `integer` (`<int>`) | Focal year for the stats in the next columns |
| `lifeExp` | `integer` (`<int>`) | Mean life expectancy in years |
| `pop` | `integer` (`<int>`) | Population size |
| `gdpPercap` | `double` (`<dbl>`) | Per-capita GDP |

: {.striped}

<hr style="height:1pt; visibility:hidden;" />

Two columns are of type `factor`,
[an alternative to](./2_vectors.qmd#factors) `character` commonly used for
categorical data.

You can also use the `View()` function to look at the data frame.
This will open a new tab in your editor pane with a spreadsheet-like look and feel:

```{r, eval=FALSE}
View(gapminder)
# (Should display the dataset in an editor pane tab)
```

::: {.callout-note appearance="simple"}
In the `head()` output above,
the `gapminder` data frame is referred to as a *"tibble"*,
which is just a slight variation on data frames from the tidyverse^[
The main difference is the nicer default printing behavior of tibbles:
e.g. the data types of columns are shown,
and only a limited number of rows are printed.].
:::

::: exercise
#### {{< fa user-edit >}} **Challenge 1: Explore the `gapminder` dataset**

TODO

:::

## `arrange()`

The `arrange()` function is like sorting functionality in Excel:
it changes the order of rows based on the values in one or more columns.
It will always rearrange the order of rows as a whole,
never just of individual columns since that would scramble the data.

`gapminder` is currently first sorted alphabetically by `country`,
and next by `year`.
But you may, for example, want to sort by population size instead,
which you can do with `arrange()` as follows:

```{r}
gm_arranged <- arrange(.data = gapminder, pop)
```

Before we take a look at the result,
let's break down the syntax of the `arrange()` function:

- The first argument is the data frame to be sorted,
  which we specify with the `.data` argument.
- The second argument is the column to sort by,
  which we specify without an argument name.

::: callout-note

Like all _dplyr_ functions, `arrange()` outputs a new data frame,
and **does not modify the input data frame.**
In this case, we chose to assign the output to a new object called `gm_arranged`.

<details><summary>[{{< fa comments >}}]{style="color: darkgreen;"} What would have happened if we had assigned the output to `gapminder`?</summary>
In that case, the original dataframe would have been overwritten with the sorted version.
</details>

<hr style="height:1pt; visibility:hidden;" />
:::

Now let's look at the result of the sorting:

```{r}
gm_arranged
```

Sorting can help you find observations with the smallest or largest values
for a certain column: above, we see that the smallest population size in
the dataset is _Sao Tome and Principe_ in 1952.

Default sorting is from **small to large**, as seen above .
To sort in reverse order, use the `desc()` (descending) helper function:

```{r}
arrange(.data = gapminder, desc(pop))
```

::: {.callout-tip appearance='simple'}
Here, we didn't assign the output to a new object, so it was just printed to screen.
We'll keep doing that for the rest of the examples in this session, 
so we can more easily see the results of the operations we are performing.
:::

Finally, you may want to sort by multiple columns,
where ties in the first column are broken by a second column (and so on).
Do this by simply listing the columns in the appropriate order:

```{r}
# Sort first by continent, then by country:
arrange(.data = gapminder, continent, country)
```

Pay attention to the syntax:
additional columns to sort by are additional arguments to the function,
not part of the same argument.
Other _dplyr_ functions work equivalently in this regard.

:::exercise
#### {{< fa user-edit >}} **Challenge 2: `arrange()`**

One observation (row) in the `gapminder` dataset has the highest per-capita GDP.
Which line of code would you use to find out which row that is?

Try it, and report which row (which country and year) you found.

::: {.panel-tabset}

#### {{< fa list-check >}} Quiz

```{r}
#| echo: false
#| results: asis
cat(longmcq(c(
  '`arrange(.data = gapminder, desc(gdpPercap), year, country)`',
  '`arrange(.data = gapminder, country, year, gdpPercap)`',
  answer = '`arrange(.data = gapminder, desc(gdpPercap))`',
  '`arrange(.data = gapminder, gdpPercap)`'
)))
```

#### {{< fa circle-info >}} Explanation

While both the first and third options will give you the row with the highest per-capita GDP,
the first option will also sort the entire dataset by year and country,
which is not necessary to answer the question.

You'll want to sort descendingly by `gdpPercap` to get the row with the highest value at the top.

```{r}
arrange(.data = gapminder, desc(gdpPercap))
```

Kuwait in 1957 had the highest per-capita GDP in the dataset, at $113,523.
:::
:::

## `filter()`

The `filter()` function outputs only those rows that satisfy one or more conditions.
It is similar to Filter functionality in Excel.

### Filter based on one condition

This first `filter()` example outputs only rows for which the
life expectancy exceeds 80 years
(remember, each row represents a country in a given year):

```{r}
filter(.data = gapminder, lifeExp > 80)
```

<details><summary>[{{< fa comments >}}]{style="color: darkgreen;"} How many rows were output?</summary>
As is printed in the first line of the output, 21 rows were output.
</details>

As the example above demonstrated,
`filter()` **outputs rows that satisfy the condition(s)** you specify.
These conditions don't have to be based on numeric comparisons:

```{r}
# Only keep rows where the value in the 'continent' column is 'Europe':
filter(.data = gapminder, continent == "Europe")
```

::: {.callout-warning appearance='simple'}
Remember to use _two_ equals signs `==` to test for equality!
:::

### Filter based on multiple conditions

It's also possible to filter based on multiple conditions.
For example, you may want to see which countries in Asia had a life expectancy
greater than 80 years:

```{r}
filter(.data = gapminder, continent == "Asia", lifeExp > 80)
```

Like above, by default, multiple conditions are combined using a Boolean ***AND***.
In other words, in a given row, _each_ condition must be met to output the row.

::: {.callout-note collapse="true"}
#### What about a Boolean ***OR***? _(Click to expand)_

If you want to combine conditions using a Boolean ***OR***,
where only one of the conditions needs to be met,
use a `|` (vertical bar) between the conditions:

```{r}
# Keep rows with a high life expectancy and/or a high GDP:
filter(.data = gapminder, lifeExp > 80 | gdpPercap > 100000)
```
:::

## The pipe (`|>`)

Examples so far applied a single _dplyr_ function to a data frame.
But in practice, it's common to use several consecutive _dplyr_ functions to
wrangle a dataframe into the format you want.

For example, you may want to filter rows and then sort the result.
You could do that as follows:

```{r}
gm_filt <- filter(.data = gapminder, lifeExp < 50, year > 2000)

arrange(.data = gm_filt, desc(pop))
```

And you could go on along these lines,
successively creating new objects that you then use for the next step.

But there is a more elegent way of dong this,
directly sending ("piping") output from one function into the next function
with the **pipe operator** `|>`
(a vertical bar followed by a greater-than sign).
Let's start by seeing a reformulation of the code above with pipes:

```{r}
gapminder |>
  filter(lifeExp < 50, year > 2000) |>
  arrange(desc(pop))
```

What happened here? We took the `gapminder` data frame,
sent ("piped") it into the `filter()` function,
whose output in turn was piped into the `arrange()` function.
You can think of the pipe as **"then"**:
take `gapminder`, _then_ filter, _then_ arrange.

When using the pipe,
you no longer specify the input data frame with the `.data` argument,
because the function now gets its input data via the pipe
(specifically, the input goes to the function's first argument by default).

::: {.callout-tip appearance='simple'}
Using pipes involves less typing and is especially more readable than using
successive assignments^[
Using pipes is also faster and uses less computer memory.
].
For code readability, 
it is good practice to always start a new line after a pipe `|>`,
and to keep the subsequent line(s) _indented_ as RStudio will automatically do.
:::

:::exercise
#### {{< fa user-edit >}} **Challenge 3: Find the mistakes**

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

This pipeline is supposed to find the 3 countries with the lowest GDP per capita
in 1997, sorted from lowest to highest.
It has two errors. Find and fix them:

```{webr}
#| autorun: false
#| caption: ""
#| exercise: find_errors
gapminder
filter(year = 1997) |>
  arrange(gdpPercap) |>
  head(n = 3)
```

```{webr}
#| exercise: find_errors
#| check: true
if (
  identical(
    .result,
    gapminder |> filter(year == 1997) |> arrange(gdpPercap) |> head(n = 3)
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa circle-check >}} Solution

**The two errors are:**

- Line 1: Missing pipe `|>` after gapminder
- Line 2: Should use `==` (not `=`) to test for equality

**Corrected code:**

```{r}
gapminder |>
  filter(year == 1997) |>
  arrange(gdpPercap) |>
  head(n = 3)
```
</details>

:::
:::

:::exercise
#### {{< fa user-edit >}} **Challenge 4: Find the extremes**

Using `filter()`, `arrange()`, and the pipe (`|>`),
answer the following questions about the gapminder dataset:

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Which African country had the highest life expectancy in 2002?

#### {{< fa circle-check >}} Solution

We'll use the same trick with `head()` like in the exercise above
to _only_ see the top 1 country, but this is not strictly necessary:
   
```{r}
gapminder |>
  filter(continent == "Africa", year == 2002) |>
  arrange(desc(lifeExp)) |>
  head(n = 1)
```

:::
::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

B) What was the smallest population recorded in Asia across all years?

#### {{< fa circle-check >}} Solution
   
```{r}
gapminder |>
  filter(continent == "Asia") |>
  arrange(pop) |>
  head(n = 1)
```
  
:::
:::

## `mutate()`

So far, we've focused on functions that subset and reorganize data frames.
But you haven't seen how you can _change the data_ or _compute derived data_.

This can be done with the `mutate()` function.
For example, to create a new column with population sizes in millions:

```{r}
# Create a new column 'pop_million' by dividing 'pop' by a million:
gapminder |>
  mutate(pop_million = pop / 10^6)
```

To _modify_ an existing column rather than adding a new one,
simply "assign back to the same name":

```{r}
# Change the unit of the 'pop' column:
gapminder |>
  mutate(pop = pop / 10^6)
```

:::exercise
#### {{< fa user-edit >}} **Challenge 5: Absolute GDP**

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

Use `mutate()` to create a new column `gdp_billion` that has the
**absolute** GDP (i.e., not relative to population size)
in **units of billions**.

```{webr}
#| autorun: false
#| caption: ""
#| exercise: abs_gdp
gapminder |>
  mutate(______)
```

```{webr}
#| exercise: abs_gdp
#| check: true
if (
  identical(
    .result,
    gapminder |> mutate(gdp_billion = gdpPercap * pop / 10^9)
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa lightbulb >}} Hints

```{r, eval=FALSE}
gapminder |>
  mutate(gdp_billion = gdpPercap * ______ / ______)
```

#### {{< fa circle-check >}} Solution

You need to do two things, which can be combined into a single line:

1. Make the GDP absolute by multiplying by the population size: `gdpPercap * pop`
2. Change the unit of the absolute GDP to billions by dividing by a billion:
   ` / 10^9`

```{r}
gapminder |>
  mutate(gdp_billion = gdpPercap * pop / 10^9)
```
</details>

:::
:::

## `summarize()`

The final function _dplyr_ function we'll cover is `summarize()`,
which computes **summaries of your data across rows**.
For example, to calculate the mean GDP across the entire dataset:

```{r}
# The syntax is similar to 'mutate': <new-column> = <operation>
gapminder |>
  summarize(mean_gdp = mean(gdpPercap))
```

The output is still a dataframe, but unlike with all previous _dplyr_ functions,
it is completely different from the input dataframe,
"collapsing" the data down to as little as a single number, like above. 

`summarize()` becomes really powerful in combination with the helper function
`group_by()` to compute **groupwise** stats.
For example, to get the mean GDP separately for each continent:

```{r}
gapminder |>
  group_by(continent) |>
  summarize(mean_gdp = mean(gdpPercap))
```

`group_by()` implicitly splits a data frame into groups of rows:
here, one group for observations from each continent.
After that, operations like in `summarize()` will happen separately for each group,
which is how we ended up with per-continent means.

:::exercise
#### {{< fa user-edit >}} **Challenge 6: Mean life expectancy**

::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

A) Calculate the average life expectancy for each country and store the result in
   a new dataframe.

```{webr}
#| autorun: false
#| caption: ""
#| exercise: life_exp
gapminder |>
  ______ |>
  ______
```

```{webr}
#| exercise: life_exp
#| check: true
if (
  identical(
    .result,
    lifeExp_bycountry <- gapminder |> group_by(country) |> summarize(mean_lifeExp = mean(lifeExp))
  )
) {
  list(correct = TRUE, message = "Nice work, that is the correct answer!")
} else {
  list(correct = FALSE, message = "That's incorrect, sorry.")
}
```

#### {{< fa lightbulb >}} Hints

```{r, eval=FALSE}
lifeExp_bycountry <- gapminder |>
  group_by(______) |>
  summarize(mean_lifeExp = ______)
```

#### {{< fa circle-check >}} Solution

First, create a dataframe with the mean life expectancy by country:

```{r}
lifeExp_bycountry <- gapminder |>
  group_by(country) |>
  summarize(mean_lifeExp = mean(lifeExp))
```

:::
::: {.panel-tabset}

#### {{< fa list-check >}} Instructions

B) Use the dataframe you just created to find out which country has the longest average
   life expectancy and which has the shortest.

#### {{< fa circle-check >}} Solution

Arrange the dataframe in two directions to see the countries with the
longest and shortest life expectancy.
You could optionally pipe into `head()` to only see the top `n`, here top 1:

```{r}
lifeExp_bycountry |>
  arrange(mean_lifeExp) |>
  head(n = 1)

lifeExp_bycountry |>
  arrange(desc(mean_lifeExp)) |>
  head(n = 1)
```

So, **Sierra Leone** has the shortest average life expectancy (36.8 years),
and **Iceland** has the longest average life expectancy (76.5 years).
:::
:::
